Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field

### 2024-05-02 paper/ms.tex
update ms
w/ good words

\received{---}
\revised{---}
\accepted{---}
\shorttitle{gyro-johannes}

@@

%
%% Early experiments for Kepler rotation samples
%\author[0000-0001-7967-1795]{Elsa~K.~Palumbo}
%\affiliation{\caltech}
%
%% ?
%%\author{Lynne~A.~Hillenbrand}
%%\affiliation{\caltech}
%
%% HIRES observations and reductions
%\author[0000-0002-0531-1073]{H. Isaacson}
%\affiliation{\berkeley}
%% HIRES observations, reductions, JUMP infrastructure
%\author[0000-0001-8638-0320]{A. W. Howard}
%\affiliation{\caltech}
%%
%\author{HIRES observers to be named}
%\affiliation{Various}

% <250 words
\begin{abstract}
  Recent studies of FGK stars in open clusters have quantitatively
  shown how stellar rotation rates and lithium content evolve over
  the first few billion years.
  %
  Here we apply this knowledge to stars observed by Kepler to compute
  rotation and lithium-based ages.
  %
  Rotation periods are drawn from previous Kepler analyses; lithium is
  measured from new and archival HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  and \nplwgyroagenograzing\ planets for which we believe
  gyrochronology is valid;
  these rotation-based ages are $\approx$15-20$\times$ more precise
  than isochrone ages in our sample of FGK stars.
  %
  The independent lithium and rotation-based ages agree for
  $\approx$90\% of stars.  However, the timescales of lithium
  depletion are such that rotation is the more precise age diagnostic
  for the majority of our stellar sample.
  %
  Our resulting catalog of planetary ages includes
  \nplyounggyrotwosigma\ planets younger than 1\,Gyr at 2$\sigma$
  (\nplyounggyrothreesigma\ planets at 3$\sigma$).
  %
  More broadly, we find that the age distribution of stars in the
  Kepler field decreases linearly from 3\,Gyr ago to present day,
  for both planet hosts as well as for the overall stellar sample.
  %
  While this stellar age distribution has implications for the
  star formation history of our Galaxy, it also bounds the relative
  balance of young and old planets detectable through
  transit surveys of nearby stars in the thin disk.
  %
\end{abstract}

\keywords{Exoplanet evolution (491), Stellar ages (1581)}

@@

Exoplanet science is in an age of abundance, fueled by the discovery
of thousands of worlds orbiting close to their host stars
\citep{Borucki10,2015JATIS...1a4003R}.  Most of these exoplanets have
ages comparable to that of the solar system.  This fact leaves many
gaps in our knowledge of the exact dynamical and physical processes
that produce these objects.  Finding and characterizing young
($<$1\,Gyr) exoplanets offers one possible workaround.  The reason is
that processes such as thermal cooling \citep{2007ApJ...659.1661F},
atmospheric loss \citep{2019AREPS..47...67O}, giant impacts
\citep{2014prpl.conf..595R}, and dynamical instability
\citep{2017MNRAS.470.1750I} are expected to over $\approx$10\,Myr to
$\approx$1\,Gyr timescales, and so they may leave observable imprints
in the properties of young planets.  Individual exemplars include
V1298~Tau \citep{David_2019}, an adolescent resonant chain of four
transiting puffy planets, and HIP~67522 \citep{Rizzuto_2020}, a
Jupiter-sized planet with a Neptunian mass (Thao et al.~submitted).
% And perhaps also TOI837, now with a uber massive core!
Population-level examples include tentative differences in the size
distribution of young planets relative to their older counterparts
\citep{Berger_2020b_rpage,David_2021,Sandoval_2021,2023AJ....166..248C,2024arXiv240303261V}.

Discovering a young planet requires solving two problems: find the
planet, and measure the star's age.  Each problem admits a range of
solutions.  Here, we will focus on planets previously found using
transits, and infer new stellar ages using rotation and lithium.
Rotation will provide the most age information for the largest number
of stars.  Lithium will provide useful information for
stars younger than a few hundred million years.

To begin, let us imagine that the age distribution of stars within a
few hundred parsecs of the Sun is uniform between 0 and 10\,Gyr
\citep[e.g.][]{Nordstrom_2004}.  This approximation is useful because
it clarifies the rarity of young stars: under these assumptions,
$\approx$1\% of nearby stars have $t$$<$100\,Myr, and $\approx$10\%
have $t$$<$1\,Gyr.  Studies of forming protoplanetary
\citep{2018A&A...617A..44K}, and of exoplanets evolving shortly after
disk dispersal \citep[e.g.][]{2022MNRAS.512.5067K}, are therefore
capped in their maximum achievable sample sizes by the tyranny of the
Galactic star formation rate.

%TODO mention the RV approach?  at least the Hyades etc work...

Despite the observational restrictions, young close-in planet
discovery has seen remarkable success over the past decade.  This has
been enabled by a combination of Kepler, K2, TESS, and Gaia
\citep[e.g.][]{Meibom_2013,Mann_K2_25_2016,Mann_2017,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Tofflemire_2021,Barber_2022,Bouma_2022a,Bouma_2022b,Zhou_2022,Zakhozhay_2022,Wood_2023}.
The original strategy from around 2010 to 2020 was to focus on stars
with known ages---obvious members of open clusters---and to search
them for transiting planets.  The resulting stellar ages are precise
at the $\approx$10\% level, and planets discovered in this way are
generally assumed to have the same age as the host cluster.  A more
recent development, facilitated by the second Gaia data release, is
the idea that finding transiting planets can often be easier than
finding and characterizing the host stellar association.

Despite recent advances, the main problem with any ``cluster-first''
approach is that only $\approx$1\% of stars within the nearest few
hundred parsecs can currently be associated with their birth cluster,
even using the Gaia data
\citep[e.g.][]{Zari_2018,CantatGaudin_2020,Kounkel_2020,Kerr_2021}.
In other words, the majority of nearby sub-gigayear stars are in the
field, and are omitted by cluster-only planet searches.  
%TODO FIXME: is this rhetorical point fair?
One can appreciate the problem by querying NASA's Exoplanet Archive
for transiting planets younger than 1\,Gyr.  Imposing reasonable
precision requirements, e.g. either $t$$<$1\,Gyr at 2$\sigma$, or else
requiring ages precise to say S/N$=$3, gives $\approx$50 ``young''
transiting planets at the time of writing.  Most of these planets are
in clusters, and have been discovered by Kepler, K2, or TESS.
However, combined, these surveys have discovered $\approx$5{,}000
planets.  Barring a major discovery bias, we would expect an order of
magnitude more transiting planets to be younger than 1\,Gyr, based on
our earlier order of magnitude reasoning.

This study aims to resolve two main issues.  First, there is the
question of how wrong it is to assume a uniform age distribution for
stars in the galactic thin disk, particularly young stars.  Second,
there is the question of which planets from the main Kepler mission
are actually young.

Given that Kepler's primary mission ended in 2013, one might be
forgiven for assuming that the ages of the Kepler planets are already
known.  Of course, they have been studied:
\citeauthor{Berger_2020b_rpage} and \citeauthor{Petigura_2022} for
instance calculated isochronal ages for individual stars, which is
most precise for stars whose luminosities and temperatures separate
them from the main-sequence.  \citeauthor{David_2021} leveraged both
isochronal ages from an earlier analogous study
\citep{Fulton_Petigura_2018_cks_vii}, and, in their core analysis,
this work used a relative gyrochrone age-ranking approach that sorted
stars based on whether they rotated faster or slower than a gyrochrone
age of $\approx$2.6\,Gyr \citep{Meibom_2015,Curtis_2020}.  These
studies were conducted in the broader context of investigating trends
in exoplanet demographics, which have been shown to depend on both
stellar metallicity and mass \citep{Petigura_2018}, at levels likely
stronger than age (at least over gigayear timescales).  However, these
studies offered limited sensitivity to the ages of the youngest
planets, due to the limitations of isochronal dating.

@@

In the planet sample, the cumulative KOI table included
\nnonfpcumkois\ confirmed and candidate planets, orbiting
\nnonfpcumkoihosts\ stars.  Assume that we apply all the same stellar
cuts, save for that on RUWE.  Then, require these confirmed and
candidate planets to have MES$>$10.  This combination of selection
criteria yield \nplwgyroagewithgrazingandhighruwe\ planets with
rotation-based ages orbiting
\nplhoststarwgyroagewithgrazingandhighruwe\ stars
(\nplhoststarwgyroagejusthighruwe\ of these stars with ``high'' RUWE).
If we additionally require the planets to be non-grazing ($b<0.8$)
based on the Markov Chain Monte Carlo (MCMC) fits from
\citet{Thompson_2018}, we are left with \nplwgyroagenograzing\ planets orbiting
\nplhoststarwgyroagenograzing\ stars.

@@

Figure~\ref{fig:li_vs_teff} shows our measured lithium equivalent
widths, plotted over the mean isochrone models from
\texttt{EAGLES} \citep{Jeffries_2023}.  We declare an upper-limit for
plotting purposes if the 1$\sigma$ lower limit on the equivalent width
is below 10\,m\AA.  Our measured
EWs span -40 to 250\,m\AA; all negative EWs have uncertainties that
are statistically consistent with zero.

@@

\paragraph{NGC~6811}
\citet{Meibom_2013} reported the discovery of Kepler-66 and -67b, two
mini-Neptunes in NGC~6811 ($\approx$1\,Gyr).  We report rotation-based
ages for these systems of \kepsixsixtgyro\ and \kepsixseventgyro\ for
Kepler-66 and -67b, respectively.  Kepler-66 ($T_{\rm
eff}$$\approx$5900\,K, $P_{\rm rot}$$\approx$10.4\,days) has a lower
precision and a more asymmetric posterior because the spindown rates
of stars this hot are relatively slow after the first gigayear.
Kepler-67 ($T_{\rm eff}$$\approx$5100\,K, $P_{\rm
rot}$$\approx$10.3\,days) is marginally hotter than the later K dwarfs
that have ``stalled'' spindown from 0.7 to 1\,Gyr, enabling its
rotation period to be diagnostic of the stellar age.

@@

% 4 CepHer, 2 MELANGE3, 6 Theia520, 2 NGC6811.
Our rotation analysis blindly recovered the 14 known Kepler planets in
clusters (Section~\ref{subsec:clusterplanets}), and provided accurate
independent ages for their host stars.

While lithium provides only marginal added information for most of
this sample, we report new robust ages from both rotation and
lithium for 14 confirmed $<$500\,Myr planets, including five
mini-Neptunes, five super-Earths, three Earths and sub-Earths, and a
Neptune-sized planet.

>>>>

\received{---}
\revised{---}
\accepted{---}
\shorttitle{Kepler's Demographic Cliff}

@@

% <250 words
\begin{abstract}
  Recent analyses of FGK stars in open clusters have revealed how
  stellar rotation rates and lithium content evolve over the first few
  billion years.
  %
  Here we apply this knowledge to stars observed by Kepler to compute
  ages using rotation and lithium.
  %
  Rotation periods are drawn from previous Kepler analyses; lithium is
  measured from new and archival HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  and \nplwgyroagenograzing\ planets for which we believe
  gyrochronology is valid; these rotation-based ages are
  $\approx$15-20$\times$ more precise than isochrone ages in our
  sample of FGK stars.
  %
  The independent lithium age estimates agree with the rotation-based
  ages for $\approx$90\% of stars.  However, the timescales of lithium
  depletion are such that rotation-based ages are more precise for
  most of our stellar sample.
  %
  Our resulting catalog of planetary ages includes
  \nplyounggyrotwosigma\ planets younger than 1\,Gyr at 2$\sigma$
  (\nplyounggyrothreesigma\ planets at 3$\sigma$).
  %
  More broadly, we find that the age distribution of stars in the
  Kepler field decreases linearly from 3\,Gyr ago to present day,
  for both planet hosts as well as for the overall stellar sample.
  %
  While this stellar age distribution has implications for the
  star formation history of our Galaxy, it also bounds the relative
  balance of young and old planets detectable through
  transit surveys of nearby stars in the thin disk.
  %
\end{abstract}

\keywords{Stellar ages (1581), Planet hosting stars (1242), Field
stars (2103), Exoplanet evolution (491), Milky Way evolution (1052),}

@@

Exoplanet science is in an age of abundance, fueled by the discovery
of thousands of worlds orbiting close to their host stars
\citep{Borucki10,2015JATIS...1a4003R}.  Most of these exoplanets are
billions of years old, similar to the Earth.  This fact leaves many
gaps in our knowledge of what exact dynamical and physical processes
produce the observed exoplanet archetypes.    The reason is that
processes such as thermal cooling \citep{2007ApJ...659.1661F},
atmospheric loss \citep{2019AREPS..47...67O}, giant impacts
\citep{2014prpl.conf..595R}, and dynamical instabilities
\citep{2017MNRAS.470.1750I} are expected to be most efficient over
$\approx$10\,Myr to $\approx$1\,Gyr timescales; when we observe most
known exoplanets, these processes are done, and not much is happening.

Finding and characterizing young ($<$1\,Gyr) exoplanets is one 
possible path toward building a timeline for exoplanet evolution.
Informative individual young systems include V1298~Tau, an adolescent
resonant chain of four transiting puffy planets that is a likely
precusor to the compact multiplanet systems \citep{David_2019}, and
HIP~67522, a Jupiter-sized planet with a Neptunian mass
(\citealt{Rizzuto_2020}; Thao et al.~submitted).
% And perhaps also TOI837, now with a uber massive core!
Population-level analyses have similarly suggested differences in the
size distribution of young exoplanets relative to their older
counterparts
\citep{Berger_2020b_rpage,David_2021,Sandoval_2021,2023AJ....166..248C,2024arXiv240303261V}.

Discovering a young planet requires solving two problems: find the
planet, and measure the star's age.  Each problem admits a range of
solutions \citep[e.g.][]{2008Sci...322.1348M,2012ApJ...756L..33Q}.  In
this article we will consider planets previously known from
transits, and infer new stellar ages using rotation and lithium.

To begin, imagine that the age distribution of stars within a few
hundred parsecs of the Sun is uniform between 0 and 10\,Gyr
\citep[e.g.][]{Nordstrom_2004}.  This approximation clarifies the
rarity of young stars: under these assumptions, $\approx$1\% of nearby
stars have $t$$<$100\,Myr, and $\approx$10\% have $t$$<$1\,Gyr.
Entire fields focused on studying forming protoplanets
\citep{2018A&A...617A..44K}, and on watching exoplanets evolve shortly
after disk dispersal \citep[e.g.][]{2022MNRAS.512.5067K}, are thereby
capped in their maximum achievable sample sizes by the tyranny of the
Galactic star formation rate.

Despite the observational challenges, young close-in planet discovery
has done well over the past decade.  This has been primarily enabled
by Kepler, K2, TESS, and Gaia
\citep[e.g.][]{Meibom_2013,Mann_K2_25_2016,Mann_2017,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Tofflemire_2021,Barber_2022,Bouma_2022a,Bouma_2022b,Zhou_2022,Zakhozhay_2022,Wood_2023}.
The original strategy from around 2010 to 2020 was to focus on stars
with known ages---obvious members of open clusters---and to search
them for transiting planets.  The resulting stellar ages are precise
at the $\approx$10\% level, and planets discovered in this way are
generally assumed to have the same age as the host cluster.  A more
recent development, facilitated by the second Gaia data release, is
the idea that finding transiting planets can often be easier than
finding and characterizing the host stellar association
\citep[e.g.][]{Tofflemire_2021}.

Despite recent advances, the main problem with the ``cluster-first''
approach is that only $\approx$1\% of stars within the nearest few
hundred parsecs are currently associated with their birth cluster
\citep[e.g.][]{Zari_2018,CantatGaudin_2020,Kounkel_2020,Kerr_2021}.
Most nearby sub-gigayear stars are in the field, and are omitted by
cluster-only planet searches.  
One can appreciate the problem by querying NASA's Exoplanet Archive
\citep[NEA;][]{2013PASP..125..989A}
for transiting planets younger than 1\,Gyr.  Imposing reasonable
precision requirements\footnote{For instance $t$$<$1\,Gyr at 2$\sigma$, or else
ages precise to say S/N$=$3.} gives $\approx$50 ``young''
transiting planets at the time of writing.  Most of these planets are
in clusters, and have been discovered by Kepler, K2, or TESS.
However, combined, those surveys have discovered $\approx$5{,}000
planets.  Unless there is a major bias against the discovery of
sub-gigayear old planets (with precisely constrained ages), we would
expect an order of magnitude more young transiting planets to be
known.

This study aims to resolve two main questions.  First, there is the
question of how wrong it is to assume a uniform age distribution for
stars in the galactic thin disk, particularly young stars.  Second,
where are the missing young transiting planets?  Briefly, we will find
that uniform is incorrect by a factor of a few, and that the missing
young transiting planets mostly have imprecise age measurements which
can be improved.

Given that Kepler's primary mission ended in 2013, one might be
forgiven for assuming that the ages of the Kepler planets are already
known.  Of course, they have been studied: \citet{Berger_2020b_rpage}
and \citet{Petigura_2022} for instance calculated isochrone ages for
Kepler Objects of Interest (KOIs).  These ages are most precise for
stars whose luminosities and temperatures separate them from the
main-sequence.  \citet{David_2021} leveraged both isochronal ages from
an earlier analogous study \citep{Fulton_Petigura_2018_cks_vii}, and
used a relative gyrochrone age-ranking approach that sorted stars
based on whether they rotated faster or slower than a gyrochrone age
of $\approx$2.6\,Gyr \citep{Meibom_2015,Curtis_2020}.  These studies
were conducted in the broader context of investigating trends in
exoplanet demographics, which also depend on stellar metallicity and
mass \citep{Petigura_2018}.  These studies offered limited sensitivity
to the ages of the youngest planets, due to the nature of isochronal
dating.

@@

In the planet sample, the cumulative KOI table included
\nnonfpcumkois\ confirmed and candidate planets, orbiting
\nnonfpcumkoihosts\ stars.  Assume that we apply all the same stellar
cuts, save for that on RUWE.  Then, require these confirmed and
candidate planets to have MES$>$10.  This combination of selection
criteria yield \nplwgyroagewithgrazingandhighruwe\ planets with
rotation-based ages orbiting
\nplhoststarwgyroagewithgrazingandhighruwe\ stars
(\nplhoststarwgyroagejusthighruwe\ of these stars with ``high'' RUWE).
If we additionally require the planets to be non-grazing ($b<0.8$),
we are left with \nplwgyroagenograzing\ planets orbiting
\nplhoststarwgyroagenograzing\ stars.

@@

Figure~\ref{fig:li_vs_teff} shows our measured equivalent widths for
the lithium 6708\AA\ absorption doublet, plotted over the mean
isochrone models from \texttt{EAGLES} \citep{Jeffries_2023}.  We
declare an upper-limit for plotting purposes if the 1$\sigma$ lower
limit on the equivalent width is below 10\,m\AA.  Our measured EWs
span -40 to 250\,m\AA; all negative EWs have uncertainties that are
statistically consistent with zero.

@@

\paragraph{NGC~6811}
\citet{Meibom_2013} reported the discovery of Kepler-66 and -67b, two
mini-Neptunes in NGC~6811 ($\approx$1\,Gyr).  We report rotation-based
ages for these systems of \kepsixsixtgyro\,Myr and \kepsixseventgyro\,Myr for
Kepler-66 and -67b, respectively.  Kepler-66 ($T_{\rm
eff}$$\approx$5900\,K, $P_{\rm rot}$$\approx$10.4\,days) has a lower
precision and a more asymmetric posterior because the spindown rates
of stars this hot slow down after the first gigayear.
Kepler-67 ($T_{\rm eff}$$\approx$5100\,K, $P_{\rm
rot}$$\approx$10.3\,days) is marginally hotter than the later K dwarfs
that have ``stalled'' spindown from 0.7 to 1\,Gyr, enabling its
rotation period to be diagnostic of the stellar age.

@@

We began this article by asking {\it i)} how wrong is the assumption
of a uniform age distribution for stars in the galactic thin disk?,
and {\it ii)} why are only $\approx$50 sub-gigayear transiting planets
known, rather than the $\approx$500 that would be expected under the 
assumption of a uniform star formation history for
the $\approx$5{,}000 known planets?

In an attempt to answer these questions, we curated a sample of rotation
periods, lithium equivalent widths, and stellar temperatures using a combination
of archival and new data.
We derived new ages using empirical interpolation-based age scales
that have recently been constructed using the latest available open cluster data.

Our analysis blindly recovered all 14 known Kepler planets in
clusters, with ages in agreement with the more precise cluster-based ages.
Although lithium provided only marginal added information for most of
the sample, the independent lithium and rotation ages agreed for
$\approx$90\% of cases.  In particular, we report new ages with strong
constraints from both rotation and lithium for 14 confirmed
$<$500\,Myr planets, including five mini-Neptunes, five super-Earths,
three Earths and sub-Earths, and a Neptune-sized planet.

With respect to our original questions however, we have 
three main conclusions.
\begin{enumerate}
  %FIXME: 120 number
  \item From the rotation rates of Kepler planet-hosting stars, we
    have identified  \nplyounggyrotwosigma\ planets younger than
    1\,Gyr at 2$\sigma$, and \nplyounggyro\ planets with median ages
    below 1\,Gyr.
    This implies that there are now $\approx$120-170 known
    sub-gigayear planets.
  %
  \item Analyzing the age distributions of both the Kepler
    target stars and the known Kepler planet-hosts, we find that
    the age-distribution of stars in the Kepler field has a
    demographic cliff.  
    There are \ratioobtoybstars\ times as many ``old'' (2-3\,Gyr)
    stars in the Kepler field as ``young'' (0-1\,Gyr) stars.
    The star formation rate today is \ratiosfr\ times
    lower than it was three billion years ago.
    %TODO: quote uncertainty on this number.
  %
  \item The implication of the latter age distribution is that rather
    than expecting $\approx$500 sub-gigayear exoplanets, we should
    instead expect $\approx$250.
    We have therefore gone from a factor of ten discrepancy to a
    factor of at most two.
    The issue is therefore mostly, although perhaps not entirely, resolved.
%    Some of the remaining missing young
%    transiting planets may be in the Kepler data.  Many more are
%    likely in K2 and TESS.
\end{enumerate}

% 4 CepHer, 2 MELANGE3, 6 Theia520, 2 NGC6811.
